---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mzRAPP

<!-- badges: start -->
<!-- badges: end -->

The goal of mzRAPP is to allow reliability assessment of non-targeted data pre-processing (NPP) in the realm of liquid chromatography high resolution mass spectrometry (LC-HRMS). This is achieved by utilizing user provided information on a set of molecules (at best > 50) with known retention behavior. mzRAPP extracts and validates chromatographic peaks for all (envipat predicted) isotopologues of those target molecules directly from mzML files. The resulting benchmark dataset is used to extract different performance measures for NPP performed on the same mzML files. 

## Installation

You can install mzRAPP from [GitHub](https://github.com/) with:

``` r
if("devtools" %in% rownames(installed.packages()) == FALSE) {install.packages("devtools")}
devtools::install_github("YasinEl/mzRAPP")
```
## Usage

Open mzRAPP as shiny app using:

```{r eval=FALSE} 
library(mzRAPP)
callmzRAPP()
```

Following that you can go to the 'Generate Benchmark' panel of mzRAPP:

<h3>Benchmark dataset generation</h3>

<h4>First:</h4>
In order to generate a benchmark you need to provide your centroided mzML files as well as two additional csv files. The <u><b>sample-group file</b></u> should contain two columns: 

<b>sample_name:</b> names of all mzML files <br>
<b>sample_group:</b> group labels of the respective samples (e.g. treated, untreated,..) <br>



The <u><b>target file</b></u> should contain information on the target molecules:

<b>molecule:</b> names of target molecules (should be unique identifiers) <br>
<b>adduct_c:</b> adducts that should be evaluated (e.g. M+H or M-Cl). If more than one adduct is to be investigated another line with the same molecule name should be added. All adducts enabled in the enviPat package are allowed:
```{r enviPat}
library(enviPat)
data(adducts)
adducts$Name
```
<b>main_adduct:</b> One main adduct has to be defined for each molecule (e.g. M+H). <br>
<b>SumForm_c:</b> molecular composition of the neutral molecule (e.g. C10H15N5O10P2). <br>
<b>StartTime.EIC:</b> Starting time for chromatograms extracted for this molecule (seconds). <br>
<b>EndTime.EIC:</b> End time for chromatograms extracted for this molecule (seconds). <br>
<b>user.rtmin:</b> (optional) Lower end of time window in which points should be considered for chromatographic peak detection (seconds). Defaults to StartTime.EIC. <br>
<b>user.rtmin:</b> (optional) Higher end of time window in which points should be considered for chromatographic peak detection (seconds). Defaults to EndTime.EIC. <br>
<b>user.rt:</b> Retention time expected for this molecule. If multiple peaks are detected the peak closest to this time is chosen (seconds). <br>

Afterwards the used instrument and resolution has to be selected. All instruments enabled via the enviPat package are enabled.


<h4>Second:</h4>
In a next step a view paramters have to be set: <br>
<b>Lowest iso. to be considered [%]:</b> Lowest relative isotopologue abundance to be considered for each molecule. <br>
<b>Min. # of scans per peak:</b> Minimum number of points for a chromatographic peak to be considered as such. <br>
<b>mz precision [ppm]:</b> Maximum spread of mass peaks in the mz dimension to be still considered part of the same chromatogram. <br>
<b>mz accuracy [ppm]:</b> Maximum difference between the accurate mz of two ion traces to be considered to be originating from the same ion. <br>
<b>Processing plan:</b> How should the benchmark generation be done? <u>sequential</u> (only using one core; often slow but does not use much RAM), <u>multiprocess</u> (using multiple cores; faster but needs more RAM; works on Windows machines) or <u>multicore</u> (using multiple cores; faster but needs more RAM; does not work on Windows machines) <br>

<h4>Third:</h4>
Benchmark generation can be started using the blue Start button. The necessary time for the generation depends on the number of mzML files, the number of target compounds and of course computational recources.

<h4>Fourth:</h4>
Afterwards the generated benchmark dataset can be inspected in the "Benchmark overview" panel. The plots in the first row can be used to inspect different qualities of the dataset. The heatmap gives an overview on which molecules were added to the benchmark (validated via at least 1 additional isotopologue). A molecule not being added does no mean that it was not there, but that mzRAPP was not able to validate it. This could be because no additional isotopologue (fullfilling strict criteria in abundance and peak shape correlation) coule be detected. The plot in the lower right corner can be used to plot peaks corresponding to individual target molecules. For additional information on the individual plot click the blue question mark-icon above the plots.

<h3>Reliability assessment of non-targeted data pre-processing</h3>

The generated benchmark can now be used to assess the reliability of non-targeted data pre-processing. Such an assessment can be set up in the panel "Assess NT data pre-processing". First the tool to be evaluated has to be set. Afterwards the unaligned files (One for XCMS (csv) and Compound Discoverer (txt), multiple for mzMine (csv) and MS-DIAL (txt)) and one aligned file have to be selected. <br>
<br>
<b>How to export unaligned and aligned files from the different tools:</b><br>
<br>
<u>XCMS (R-version):</u>
```{r eval=FALSE} 
#unaligned file:
data.table::fwrite(xcms::peaks(xcmsSet_object), "blabla_unaligned_file.csv")

#aligned file:
data.table::fwrite(xcms::peakTable(xcmsSet_object), "blabla_aligned_file.csv")
```
<br>
<u>XCMS online:</u> <br>
Download all results from XCMS online via the button "Download Results". Afterwards extract all Results from the zipped folder. <br>
unaligned file: select the xcms3xset.Rda file <br>
aligned file: select the same xcms3xset.Rda file <br>
<br>
<u>MS-DIAL:</u> <br>
unaligned files: Export -> Peak list result -> [Add all files] -> [set Export format to txt] <br>
aligned file: Export -> Alignment result -> [check Raw data matrix Area] -> [set Export format to txt] <br>
<br>
<u>mzMine:</u> <br>
unaligned files: [select all files generated in the chromatogram deconvolution step] -> Feature list methods -> Export/Import -> Export to CSV file -> [set Filename including pattern/curly brackets (e.g. blabla_{}_blabla.csv)] -> [check "Peak height", "Peak area", "Peak RT start", "Peak RT end", "Peak RT", "Peak m/z", "Peak m/z min" and "Peak m/z max"] -> [set Filter rows to ALL] <br>
aligned file: [select file after alignment step] -> Feature list methods -> Export/Import -> Export to CSV file -> [additional to checks set for unaligned files check "Export row retention time" and "Export row m/z"]<br>
<br>
<u>Compound Discoverer:</u> <br>
unaligned file: [go to panel "Features"] -> [click in any row] -> [click "Show related Tables" on the bottom of the screen] -> [go to panel "Chromatogram Peaks"] -> [make sure the columns "Apex Intensity", "Area", "Study File ID", "Left RT [min]", "Right RT [min]" "Apex RT [min]" and "Apex m/z" are visible] -> [right click any row] -> Export -> As plain text... <br>
<br>
<b>Selecting a benchmark dataset and starting assessment:</b><br>
<br>
Next the benchmark file has to be selected. If a benchmark has been created during this shiny session (the benchmark is still visible in the panel benchmark overview) the switch button "Use generated benchmark" can be clicked as an alternative. If you want to adapt mzRAPP to a tool/non-targeted output format other than those mentioned above this can be done by clicking the "Use generated options" switch button. The way those option files can be generated manually is described at the bottom of this readme. <br>
<br>
After performing those steps the assessment can be started via the blue "Start assessment button".

```{r pressure, echo = FALSE}
#shiny::runApp("callmzRAPP", display.mode = "showcase")
```


## Interpretation of assessment results

Reliability assessment results are distributed over the panels "Assessment results (peaks)" and "Assessment results (alignment)". In the first panel a number of performance metrics is given on the top: <br>
<br>
<u>Found peaks:</u> <br>
The number of benchmark peaks for which a match was found among the NT peaks in the peak picking step. <br>
<br>
<u>Peak fragments:</u> <br>
The number of peak fragments which have been found that have been found over the whole dataset. For a graphical explanation of a split peak please look at the mzRAPP publication. <br>
<br>
<u>Missing Value (S|R):</u> <br>
Number of systematic (s) and random (R) missing values after the peak picking step. This classification is depending on the peak with the lowest benchmark area detected via NPP. If a missing peak has a benchmark area > 20% higher than the lowest NPP-detected peak it is classified as R if it is < than 20% as S. <br>
<br>
<u>Pred. error increase >20%p:</u> <br>
Isotopologue areas are predicted from the most abundant isotopologue of each compound. If the relative error of this prediction is more than 20 %p higher for NT-peak areas (compared to benchmark-areas) it is refelcted in this variable. Here only peak areas from the peak picking step are considered. <br>
<br>
<u>Min. # of alignment errors:</u> <br>
Gives the minimum number of alignment errors done by the non-targeted algorithm. For the exact way of how this count is esatblished please check the original mzRAPP publication. <br>
<br>
<u>Found peaks (features):</u> <br>
The number of benchmark peaks for which a match was found among the NT peaks after the alignment step. Here only one feature per benchmark feature is allowed. If multiple features are found for one benchmark feature the one which fits the predicted isotopic pattern best is chosen. <br>
<br>
<u>Found features (features):</u> <br>
The number of benchmark features for which at least one benchmark peak was detected via the non-targeted algorithm. <br>
<br>
<u>Pred. error increase >20%p (features):</u> <br>
Isotopologue areas are predicted from the most abundant isotopologue of each compound. If the relative error of this prediction is more than 20 %p higher for NT-peak areas (compared to benchmark-areas) it is refelcted in this variable. Here only peak areas after the alignment step are considered. <br>
<br>
<b>Overview plots:</b><br>
<br>
For explanations of the overview plots please click the blue question marks above the individual plots.




